---
title: "**results: Amsterdam subgraphs**"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
require(ggplot2)
require(ggpubr)
require(gridExtra)
require(tidyr)
```
```{r, include=FALSE, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(rmarkdown)
setwd('~/Documents/GitHub/hivclust/pkg/misc/vignettes') 
rmarkdown::render('roadmap.200203.phyloscanner.subgraphs.Rmd')
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
read_chunk('~/Documents/GitHub/hivclust/pkg/misc/hivclu.ROADMAP.R')		
```

## **Extracting phylogenetic subgraphs associated with viral spread in Amsterdam**
We start by extracting the phylogenetic subgraphs that are associated with viral
spread in Amsterdam. If we interpret each subgraph
as a distinct transmission chain in Amsterdam, (which may
be partially observed due to incomplete sequence sampling), 
we can explore patterns within transmission chains. 
```{r,include=TRUE, eval=FALSE, echo=TRUE, warning=FALSE,message=FALSE, tidy=FALSE} 
<<rmd.chunk.roadmap.200203.get.dated.subgraphs>>
```

Just a few notes: a large fraction of the subgraphs consist of just one sequence. 
These subgraphs don't contain information into viral spread within Amsterdam 
because no one else of the corresponding transmission chain was sampled, 
and so we can ignore them in a phylodynamic analysis. This reduces sample size, 
and makes the analysis more tractable.

On the other hand, the subgraphs of size 1 also provide useful information into
the reproductive number, because they indicate stuttering viral spread, and a
reproduction number below 1. So we don't want to exclude them when estimating
reproductive numbers. 

## **Reading phylogenetic subgraphs **
First we read in the taxa that are part of Amsterdam phylogenetic subgraphs
into R: 
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE,tidy=FALSE} 
require(data.table)
require(phangorn)
require(ggplot2)
require(reshape)

#	working directory with the dated subgraphs		
#home <- '~/Documents/Roadmap/Data/SHM_1902_ROADMAP_191223'
home <- '~/Box Sync/Roadmap/analysis_200407'
indir.phsc	<- file.path(home,'subgraphs_dated')

#	extract subgraph taxa	
infiles <- data.table(F=list.files(indir.phsc, pattern='rda$', full.names=TRUE, recursive=TRUE))	
infiles <- subset(infiles, grepl('datedsubgraphs',F))
infiles <- subset(infiles, grepl('datedsubgraphs',F))
#	SELECT defines if Ams, AmsHSX, AmsMSM
infiles[, SELECT:= gsub('^(.*)subgraphs_([A-Za-z]+).rda','\\2',basename(F))]
#	for subtype B, I ran separate analyses based on very large subtrees for comp efficiency
#	ST stores the subtype, and ST_CLADE the large subtree
infiles[, ST:= gsub('^.*subtype_([^_]+)_.*\\.rda','\\1',basename(F))]
# For B only
infiles[, ST_CLADE:= as.integer(gsub('[^c]*c([0-9]+)','\\1',ST))]
infiles[, ST:= gsub('([^c]*)c([0-9]+)','\\1',ST)]
#	ID of bootstrap replicate, 000 for analysis on real alignment
infiles[, REP:= gsub('^.*wOutgroup_([0-9]+)_.*\\.rda','\\1',basename(F))]
dsubgraphtaxa <- infiles[, {
  #i<- 1
  #infile <- infiles[i, F]			
  infile <- F
  cat('Process',infile,'\n')
  load(infile)			
    if(length(subgraphs)==1){
        subgraph.names <- rep(subgraphs[[1]]$subgraph.name, length(subgraphs[[1]]$tip.label))
        subgraph.taxa <- subgraphs[[1]]$tip.label			
  }  else{
    subgraph.names <- unlist(sapply( subgraphs, function(subgraph)  rep(subgraph$subgraph.name, length(subgraph$tip.label)) ))
    subgraph.taxa <- unlist(sapply( subgraphs, function(subgraph)  subgraph$tip.label))
  }
  list(NAME=subgraph.names, 
      TAXA= subgraph.taxa 
  )				
}, by=c('ST','ST_CLADE','REP','SELECT')]	
#	add meta data from taxa names
regex.tip.label <- '^([A-Za-z]+)_+([0-9]+)_([0-9]+)_([a-zA-Z]+)_([a-zA-Z]+)_([a-zA-Z]+)-([a-zA-Z]+)_([a-zA-Z_]+)_([0-9]+)-([0-9-]+)-([0-9-]+)_([0-9]+)$'
dsubgraphtaxa[, ID:= as.numeric(gsub(regex.tip.label,'\\3',TAXA))]	
dsubgraphtaxa[, SEQ_YEAR:= as.numeric(gsub(regex.tip.label,'\\9',TAXA))]
dsubgraphtaxa[, FULL_NAME:= paste0(ST,'_',REP,'_',SELECT,'_',NAME)]
```

## **Adding meta-data **
Next, I add meta-data on the individuals:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE}
infile.meta <- file.path(home,'misc','200327_sequence_labels.rda')

#	add meta data from pre-processed baseline data	
load(infile.meta)
dind <- as.data.table(dind)	
setnames(dind, c('PATIENT','BIRTH_Y','BIRTH_CNTRY'), c('ID','BIRTH_YEAR','BIRTH_COUNTRY'))
tmp <- subset(dind, select=c(ID,BIRTH_YEAR,BIRTH_COUNTRY,LOC_BIRTH,CITY,TRANSM,GENDER))
set(tmp, NULL, 'BIRTH_YEAR', tmp[, as.numeric(BIRTH_YEAR)])
dsubgraphtaxa <- merge(dsubgraphtaxa,tmp,by='ID')
dsubgraphtaxa[, AGE2020:= cut(2020-BIRTH_YEAR, breaks=c(-Inf,20,25,30,35,40,45,50,55,60,Inf))]
```


## **Distribution of subgraph sizes**
Let us first have a look on the distribution of subgraph sizes, that is how
many individuals there are in partially observed, separate transmission chains.
(The transmission chains are partially observed because we do not have
sequence data from everyone).

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 4), fig.align="center"}
dsubgraphsize <- dsubgraphtaxa[, list(SIZE=length(ID)), by=c('ST','REP','SELECT','NAME')]
dsubgraphsize[, SINGLETON:= as.character(factor(SIZE==1, levels=c(TRUE,FALSE), labels=c('subgraph size 1','subgraph size >1')))]
ggplot(subset(dsubgraphsize, SELECT!='Ams'), aes(x=SIZE, fill=ST)) + 
		geom_bar() + 
		facet_grid(SELECT~.) +
		labs(x='\nsize of phylogenetic subgraphs attributed to Amsterdam')

```

## **Sample size for phylodynamic analysis of HIV spread in Amsterdam **
Next, let us have a look at how individuals are in subtrees are of size 1 and
size >1.

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 4), fig.align="center"} 
#	number of individuals in subgraphs of size == 1 or size > 1 
dsubgraphind <- dsubgraphsize[, list(N_INDIVIDUALS=sum(SIZE)), by=c('REP','ST','SELECT','SINGLETON')]
dsubgraphind <- dcast.data.table(dsubgraphind, REP+ST+SINGLETON~SELECT,value.var='N_INDIVIDUALS') 
dsubgraphind[1:14]
```
	
The number of subgraphs of size >1 differ by subtype according to the trasnsmission group (HSX/MSM).
 For example, subtypes 02_AG and C have the most subgraphs of size >1 amongst HSX, but 
 subtype 01_AE has many more subgraphs of size >1 amongst MSM.

## **Empirical analysis on group-mixing within subgraphs **
	
To quantify the extent of mixing across certain population groups, 
we first select just those subgraphs of size >1:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
#	number of individuals in subgraphs of size == 1 or size > 1 
tmp <- subset(dsubgraphsize, SIZE>1, c(ST, REP, SELECT, NAME))
dsubgraphtaxa2 <- merge(dsubgraphtaxa, tmp, by=c('ST','REP','SELECT','NAME'))
```
   
### **Place of birth **

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2[, DUMMY:= as.character(factor(BIRTH_COUNTRY=='Netherlands',levels=c(TRUE,FALSE),labels=c('Dutch-born','other')))]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='Ams'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
  geom_bar(stat='identity') + 
  labs(y='#individuals',x='phylo subgraphs',fill='birth place') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  facet_grid(SELECT~ST, scale='free_x', space='free_x')
```

To make things a bit clearer we can also plot the number of subgraphs which are composed of >=75% Dutch-born individuals, >=75% of individuals born outside of the Netherlands and otherwise mixed:
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','REP','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','REP','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','REP','SELECT','ST'))
dcnt$PROP <- dcnt$N/dcnt$TOTAL
bp <- dcnt[, list(N75= length(FULL_NAME[PROP>=0.75])), by=c('SELECT','REP','ST','DUMMY')]
nsubgraphs <- dsubgraphtaxa2[, list(SIZE=length(unique(NAME))), by=c('REP','ST','SELECT')]
bp <- merge(bp,nsubgraphs, by=c('REP','ST','SELECT'))
mixed <- bp[, list(Mixed= SIZE-sum(N75)), by=c('REP','ST','SELECT')]
bp <- spread(bp, DUMMY, N75)
bp <- merge(bp,mixed,by=c('REP','ST','SELECT'))
bp <- unique(bp)
bp <- melt(bp, id.vars=c('REP', 'ST', 'SELECT', 'SIZE'))
bp$bpp <- factor(bp$variable, levels=c('Dutch-born','other','Mixed'),labels=c(">75% NL", ">75% MW-MB", "Mixed"))
nonbl <- ggplot(subset(bp, SELECT!='Ams' & ST!='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='#subgraphs',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank()) +
  facet_grid(SELECT~., scale='free_x', space='free_x')
legend <- get_legend(nonbl)
nonb <- ggplot(subset(bp, SELECT!='Ams' & ST!='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='#subgraphs',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank(),legend.position="none") +
  facet_grid(SELECT~., scale='free_x', space='free_x')
b <- ggplot(subset(bp, SELECT!='Ams' & ST=='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank(),legend.position="none") +
  facet_grid(SELECT~., scale='free_x', space='free_x')
grid.arrange(nonb, b,
                      legend,
                      ncol = 3, widths=c(3.5, 1.5, 1.5)) 
```


```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```
 
The results suggest that

* Amongst HSX, there appear to be more indivuduals born outside of the NL than Dutch-born, with not much mixing, and subgraphs are generally quite small (many of size 2). There appears to be more mixing of groups for subtypes C and D, but subgraphs are still small (size 2/3)
* There are many more Dutch-born individuals in the subgraphs for MSM than foreign-born individuals, with some mixing

### **Place of birth, geographic region **
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2$LOC_BIRTH[dsubgraphtaxa2$BIRTH_COUNTRY=='Netherlands'] <- dsubgraphtaxa2$BIRTH_COUNTRY[dsubgraphtaxa2$BIRTH_COUNTRY=='Netherlands']
dsubgraphtaxa2[, DUMMY:= factor(LOC_BIRTH, levels=c('Netherlands','WEurope','EEurope','Africa','Asia','LaAmCarr','NorthAm','Oceania','Unknown'),labels=c('Netherlands','WEurope','EEurope','Africa','Asia','LaAmCarr','NorthAm','Oceania','Unknown'))]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='Ams'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
  geom_bar(stat='identity') + 
  labs(y='#individuals',x='phylo subgraphs',fill='birth place') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  facet_grid(SELECT~ST, scale='free_x', space='free_x')

dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','REP','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','REP','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','REP','SELECT','ST'))
dcnt$PROP <- dcnt$N/dcnt$TOTAL
bp <- dcnt[, list(N75= length(FULL_NAME[PROP>=0.75])), by=c('SELECT','REP','ST','DUMMY')]
nsubgraphs <- dsubgraphtaxa2[, list(SIZE=length(unique(NAME))), by=c('REP','ST','SELECT')]
bp <- merge(bp,nsubgraphs, by=c('REP','ST','SELECT'))
mixed <- bp[, list(Mixed= SIZE-sum(N75)), by=c('REP','ST','SELECT')]
bp <- spread(bp, DUMMY, N75)
bp <- merge(bp,mixed,by=c('REP','ST','SELECT'))
bp <- unique(bp)
bp <- melt(bp, id.vars=c('REP', 'ST', 'SELECT', 'SIZE'))

bp$bpp <- factor(bp$variable, levels=c('Netherlands','WEurope','EEurope','Africa','Asia','LaAmCarr','NorthAm','Oceania','Unknown','Mixed'),labels=c('>75% Netherlands','>75% WEurope','>75% EEurope','>75% Africa','>75% Asia','>75% LaAmCarr','>75% NorthAm','>75% Oceania','>75% Unknown','Mixed'))
bp$bpc <- bp$variable
levels(bp$bpc) <- c(levels(bp$bpc),"Netherlands+FormerCurrTerritories")
bp$bpc[bp$variable=="Netherlands" | bp$variable=="FormerCurrDutchColonies"] <- 'Netherlands+FormerCurrTerritories'

nonbl <-  ggplot(subset(bp, SELECT!='Ams' & ST!='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='#subgraphs',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank()) +
  facet_grid(SELECT~., scale='free_x', space='free_x')
legend <- get_legend(nonbl)
nonb <- ggplot(subset(bp, SELECT!='Ams' & ST!='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='#subgraphs',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank(),legend.position="none") +
  facet_grid(SELECT~., scale='free_x', space='free_x')
b <-   ggplot(subset(bp, SELECT!='Ams' & ST=='B'), aes(x=ST, y=value, fill=bpp)) +
  geom_bar(stat='identity') + 
  labs(y='',x='subtype',fill='birth place') +
  theme(axis.ticks.x=element_blank(),legend.position="none") +
  facet_grid(SELECT~., scale='free_x', space='free_x')
grid.arrange(nonb, b,
                           legend,
                           ncol = 3, widths=c(3.5, 1.5, 1.5)) 
```

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```


* There appears to be much more evidence of mixing of birth regions amongst MSM than HSX, and subgraphs are generally larger for
subtypes 01_AE, 01_AG, A1
* There don't appear to be many subgraphs predominantly composed of individuals from Morroco or Turkey, suggesting dissatortative mixing within chains including these sequences

### **Age group **
Next, we consider age brackets for comparison:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2[, DUMMY:= AGE2020]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='Ams'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
  geom_bar(stat='identity') + 
  labs(y='#individuals',x='phylo subgraphs',fill='age in 2020') +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  facet_grid(SELECT~ST, scale='free_x', space='free_x')
```

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```
 
There are a lot more groups to compare, but some things I noticed:

* The degree of mixing, and the size of the subgraphs does seem to differ by subtype.
For example, subtype 01_AE suggests some mixing between age groups of MSM, but amongst HSX 
the subgraphs are small (mostly size 2) and there is less mixing. Also, subtypes D and 
06_cpx do not have any subgraphs > size 1
