---
title: "**results: King County subgraphs**"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
require(ggplot2)
```
```{r, include=FALSE, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(rmarkdown)
setwd('~/git/hivclust/pkg/misc/vignettes') 
rmarkdown::render('seattle.191017.phyloscanner.subgraphs.Rmd')
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
#source('~/git/hivclust/pkg/misc/hivclu.Seattle.R')
read_chunk('~/git/hivclust/pkg/misc/hivclu.Seattle.R')		
```

## **Reading phylogenetic subgraphs **
I first read the taxa that are part of King County phylogenetic subgraphs: 
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
require(data.table)
require(phangorn)
require(ggplot2)
require(reshape)

#	working directory with phyloscanner output		
home <- '~/Box Sync/OR_Work/Seattle'
indir.phsc	<- file.path(home,'analysis_191017','phyloscanner')

#	extract subgraph taxa	
infiles <- data.table(F=list.files(indir.phsc, pattern='rda$', full.names=TRUE, recursive=TRUE))	
infiles <- subset(infiles, grepl('datedsubgraphs',F))
#	SELECT defines if KC, KCHSX, KCMSM
infiles[, SELECT:= gsub('^(.*)subgraphs_([A-Z]+)\\.rda','\\2',basename(F))]
#	for subtype B, I ran separate analyses based on very large subtrees for comp efficiency
#	ST stores the subtype, and ST_CLADE the large subtree
infiles[, ST:= gsub('^.*Subtype([^_]+)_.*\\.rda','\\1',basename(F))]
infiles[, ST_CLADE:= as.integer(gsub('[^c]*c([0-9]+)','\\1',ST))]
infiles[, ST:= gsub('([^c]*)c([0-9]+)','\\1',ST)]
#	ID of bootstrap replicate, 000 for analysis on real alignment
infiles[, REP:= gsub('^.*ndrm_([0-9]+)_.*\\.rda','\\1',basename(F))]
dsubgraphtaxa <- infiles[, {
			#i<- 1
			#infile <- infiles[i, F]			
			infile <- F
			cat('Process',infile,'\n')
			load(infile)			
			subgraph.names <- unlist(sapply( subgraphs, function(subgraph)  rep(subgraph$subgraph.name, length(subgraph$tip.label)) ))
			subgraph.taxa <- unlist(sapply( subgraphs, function(subgraph)  subgraph$tip.label))				
			list(	NAME=subgraph.names, 
					TAXA= subgraph.taxa 
			)				
		}, by=c('ST','ST_CLADE','REP','SELECT')]	
#	add meta data from taxa names
regex.tip.label <- '^([A-Z]+)_+PR/RT-([0-9]+)_([0-9]+)_([a-zA-Z]+)_([a-zA-Z]+)_([a-zA-Z]+)_([a-zA-Z]+)_([a-zA-Z_]+)_([0-9]+)$'
dsubgraphtaxa[, ID:= as.numeric(gsub(regex.tip.label,'\\3',TAXA))]	
dsubgraphtaxa[, SEQ_YEAR:= as.numeric(gsub(regex.tip.label,'\\9',TAXA))]
dsubgraphtaxa[, FULL_NAME:= paste0(ST,'_',REP,'_',SELECT,'_',NAME)]
```

## **Adding meta-data **
Next, I add meta-data on the individuals:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE}
infile.meta <- file.path(home,'analysis_191017','misc','180709_sequence_labels.rda')

#	add meta data from pre-processed persons file	
load(infile.meta)
dind <- as.data.table(dind)	
setnames(dind, c('newnum','b_yr','birthCountry2','county','race2','transm2','Gender2'), c('ID','BIRTH_YEAR2','BIRTH_COUNTRY2','COUNTY','RACE2','TRANSM2','GENDER2'))
tmp <- subset(dind, select=c(ID,BIRTH_YEAR2,BIRTH_COUNTRY2,COUNTY,RACE2,TRANSM2,GENDER2))
set(tmp, NULL, 'BIRTH_YEAR2', tmp[, as.numeric(BIRTH_YEAR2)])
dsubgraphtaxa <- merge(dsubgraphtaxa,tmp,by='ID')
dsubgraphtaxa[, AGE2020:= cut(2020-BIRTH_YEAR2, breaks=c(-Inf,20,25,30,35,40,45,50,55,60,Inf))]
```


## **Distribution of subgraph sizes**
Let us first have a look on the distribution of subgraph sizes, that is how
many individuals there are in partially observed, separate transmission chains.
(The transmission chains are partially observed because we do not have
sequence data from everyone).

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 4), fig.align="center"}
dsubgraphsize <- dsubgraphtaxa[, list(SIZE=length(ID)), by=c('ST','REP','SELECT','NAME')]
dsubgraphsize[, SINGLETON:= as.character(factor(SIZE==1, levels=c(TRUE,FALSE), labels=c('subgraph size 1','subgraph size >1')))]
ggplot(subset(dsubgraphsize, SELECT!='KC'), aes(x=SIZE, fill=ST)) + 
		geom_bar() + 
		facet_grid(SELECT~.) +
		labs(x='\nsize of phylogenetic subgraphs attributed to King County/Seattle')
```

## **Sample size for phylodynamic analysis of HIV spread in Seattle **
Next, let us have a look at how individuals are in subtrees are of size 1 and
size >1. The number of individuals in subgraphs of size > 1 can be informative
about transmission dynamics within Seattle. 

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 4), fig.align="center"} 
#	number of individuals in subgraphs of size == 1 or size > 1 
dsubgraphind <- dsubgraphsize[, list(N_INDIVIDUALS=sum(SIZE)), by=c('ST','REP','SELECT','SINGLETON')]
dsubgraphind <- dcast.data.table(dsubgraphind, ST+REP+SINGLETON~SELECT,value.var='N_INDIVIDUALS') 
dsubgraphind
```
	
There are a little more than 350 individuals in HSX subgraphs of size > 1. That
s not too bad! There are more than 4000 individuals in MSM subgraphs of size >
1. That s a lot and I am not sure how to run a phylodynamic model on this sample
   size.

## **Empirical analysis on group-mixing within subgraphs **
	
Next, I was interested to quantify the extent of mixing across certain
population groups, based on the subgraphs of size > 1. So let us select just
those:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
#	number of individuals in subgraphs of size == 1 or size > 1 
tmp <- subset(dsubgraphsize, SIZE>1, c(ST, REP, SELECT, NAME))
dsubgraphtaxa2 <- merge(dsubgraphtaxa, tmp, by=c('ST','REP','SELECT','NAME'))
```

### **Ethnicity **
First, I considered ethnicity.

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2[, DUMMY:= RACE2]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='KC'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
		geom_bar(stat='identity') + 
		labs(y='#individuals',x='phylo subgraphs',fill='ethnicity') +
		theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
		facet_grid(SELECT~ST, scale='free_x', space='free_x')
```
We previously found for non-B subtypes that the subgraphs were homogeneous by ethnicity.
However, the large majority of patients are subtype B, and it now seems that there is considerable mixing
by ethnicity. 

Let us quantify this. As a measure of mixing, I quantified average homogeneity
of each subgraph [as described here](https://stats.stackexchange.com/questions/221332/variance-of-a-distribution-of-multi-level-categorical-data), as well as the average of the largest
frequency of any strata: 
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```
 
The results suggest that

* there are a large number of separate transmission chains in King County 
* they are typically small for heterosexuals
* each of them typically contains individuals of more than one ethnic group.
  On average, individuals of white ethnicity make up around 65% of individuals
  in a transmission chain, both for MSM and HSX spread.
   
### **Place of birth **
Second, I considered place of birth.

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2[, DUMMY:= as.character(factor(BIRTH_COUNTRY2=='USA',levels=c(TRUE,FALSE),labels=c('US-born','other')))]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='KC'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
		geom_bar(stat='identity') + 
		labs(y='#individuals',x='phylo subgraphs',fill='birth place') +
		theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
		facet_grid(SELECT~ST, scale='free_x', space='free_x')
```

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```
 
The results suggest that

* mixing is more pronounced in terms of ethnicity compared to place of birth.

### **Age group **
Third, I considered age brackets for comparison.

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE, fig.dim = c(8, 6), fig.align="center"} 
dsubgraphtaxa2[, DUMMY:= AGE2020]
dcnt <- dsubgraphtaxa2[, list(N= length(ID)), by=c('FULL_NAME','SELECT','ST','DUMMY')]
tmp <- dsubgraphtaxa2[, list(TOTAL= length(ID)), by=c('FULL_NAME','SELECT','ST')]
dcnt <- merge(dcnt, tmp, by=c('FULL_NAME','SELECT','ST'))
ggplot(subset(dcnt, SELECT!='KC'), aes(x=FULL_NAME, y=N, fill=DUMMY)) +
		geom_bar(stat='identity') + 
		labs(y='#individuals',x='phylo subgraphs',fill='age in 2020') +
		theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
		facet_grid(SELECT~ST, scale='free_x', space='free_x')
```

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=FALSE} 
tmp <- dcnt[, list(R= sum( (N/TOTAL)^2), PMAX= max(N/TOTAL) ), by=c('FULL_NAME','SELECT','ST')]
tmp[, list(R_AVG=mean(R), PMAX_AVG=mean(PMAX)), by=c('SELECT')]
```
 
I am not clear how to interpret the results in comparison to ethnicity or place
of birth, because there are many more age groups.