---
title: "**Phylodynamic analysis: transmission dynamics between US born and foreign born individuals in King County Seattle**"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
require(ggplot2)
```
```{r, include=FALSE, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(rmarkdown)
setwd('~/git/hivclust/pkg/misc/vignettes') 
rmarkdown::render('seattle.191017.phydy.model.Rmd')
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
#source('~/git/hivclust/pkg/misc/hivclu.Seattle.R')
read_chunk('~/git/hivclust/pkg/misc/hivclu.Seattle.R')		
```

## **Epidemiologic model, local population**
The overall aim is to specify an epidemic model that captures transmission
dynamics among heterosexuals by migrant-status. Consider the following states
$$
(S^j_{gi}, I^j_{gi}, T^j_{gi})
$$ 
where $S$, $I$, $T$ denote respectively groups of susceptible, infected and
treated individuals and 

* $j$ denotes location, respectively $l$ (local, King County) or $x$
(external, world) 
* $g$ denotes gender, respectively $m$ (male) and $j$
(female) 
* $i$ denotes migration status. This can be ethnicity, respectively
$0$ (white) and $1$ (non-white). Or alternatively location of birth,
respectively $0$ (US-born) and $1$ (foreign-born). 

Our basic ODE equations for the transmission dynamics involving the local
susceptible population are 
$$
\begin{align}
    \dot{S}^l_{f0} &= - (\beta_{00} I^l_{m0}+\beta_{10} I^l_{m1} ) S^l_{f0} /N^l + \mu (S^l_{f0}/S^l) - \mu S^l_{f0} \\
    \dot{S}^l_{f1} &= - (\beta_{01} I^l_{m0}+\beta_{11} I^l_{m1} ) S^l_{f1} /N^l + \mu (S^l_{f1}/S^l) - \mu S^l_{f1}\\
    \dot{S}^l_{m0} &= - (\beta_{00} I^l_{f0}+\beta_{10} I^l_{f1} ) S^l_{m0} /N^l + \mu (S^l_{m0}/S^l) - \mu S^l_{m0} \\
    \dot{S}^l_{m1} &= - (\beta_{01} I^l_{f0}+\beta_{11} I^l_{f1} ) S^l_{m1} /N^l + \mu (S^l_{m1}/S^l) - \mu S^l_{m1}\\ 
\end{align}
$$
where 

* $\mu$ is the birth/death rate
* $S^l = S^l_{f0} + S^l_{f1} + S^l_{m0} + S^l_{m1}$
* $N^l$ is the sum of all compartments involving the local population
* $\beta_{ij}$ is the transmission rate from individuals with migration status $i$ to migration status $j$. We are primarily interested in the relative differences between these rates,
and so we write $\beta_{ij}= \beta * b_{ij}$ and constrain $\sum_{i,j} b_{ij} = 0$.

The remaining model equations involving the local population are  
$$
\begin{align}
    \dot{I}^l_{f0} &= (\beta_{00} I^l_{m0}+\beta_{10} I^l_{m1} ) S^l_{f0} /N^l - \gamma I^l_{f0} - \mu I^l_{f0} \\
    \dot{I}^l_{f1} &= (\beta_{01} I^l_{m0}+\beta_{11} I^l_{m1} ) S^l_{f1} /N^l - \gamma I^l_{f1} - \mu I^l_{f1}\\
    \dot{I}^l_{m0} &= (\beta_{00} I^l_{f0}+\beta_{10} I^l_{f1} ) S^l_{m0} /N^l - \gamma I^l_{m0} - \mu I^l_{m0} \\
    \dot{I}^l_{m1} &= (\beta_{01} I^l_{f0}+\beta_{11} I^l_{f1} ) S^l_{m1} /N^l - \gamma I^l_{m1} - \mu I^l_{m1}\\
    \dot{T}^l_{f0} &= \gamma I^l_{f0} - \mu T^l_{f0} \\
    \dot{T}^l_{f1} &= \gamma I^l_{f1} - \mu T^l_{f1}\\
    \dot{T}^l_{m0} &= \gamma I^l_{m0} - \mu T^l_{m0} \\
    \dot{T}^l_{m1} &= \gamma I^l_{m1} - \mu T^l_{m1}\\ 
\end{align}
$$
where

* $\gamma$ is the rate at which infected individuals become virally suppressed due to
antiretrovirals. We assume that infected individuals that reached viral
suppression remain suppressed until death.

## **Epidemiologic model, connection with external world**
Considering that the local transmission dynamics do not occur in isolation of
the rest of the world, we make the following assumptions

* Local individuals with migration status $0$ can acquire infections from external infecteds at rate $\beta_{20}$.
* Local individuals with migration status $1$ can acquire infections from external infecteds at rate $\beta_{21}$.  
* External individuals can in-migrate to the group with migration status $1$ at rate $\nu$.
* Local individuals do not out-migrate, and neither do they infect external
  susceptibles. This is to keep the model simple, and also reflects our data
  (sequence samling fraction of Seattle epidemic much larger than sequence
  samling fraction of external epidemic) 

The ODE equations involving the external population are 
$$
\begin{align}   
    \dot{S}^x_{f} &= - \xi I^x_{m} S^x_{f} /N^x + \mu (S^x_{f}/S^x) - \mu S^x_{f} -\nu S^x_{f}\\
    \dot{S}^x_{m} &= - \xi I^x_{f} S^x_{m} /N^x + \mu (S^x_{m}/S^x) - \mu S^x_{m} -\nu S^x_{m}\\
    \dot{I}^x_{f} &= \xi I^x_{m} S^x_{f} /N^x - \zeta I^x_{f} - \mu I^x_{f} - \nu I^x_{f}\\
    \dot{I}^x_{m} &= \xi I^x_{f} S^x_{m} /N^x - \zeta I^x_{m} - \mu I^x_{m} - \nu I^x_{m}\\
    \dot{T}^x_{f} &= \zeta I^x_{f} - \mu T^x_{f} - \nu T^x_{f}\\
    \dot{T}^x_{m} &= \zeta I^x_{m} - \mu T^x_{m} - \nu T^x_{m}, 
\end{align}
$$
with a separate transmission rate $\xi$, a separate suppression rate $\zeta$ and the same birth/death rate $\mu$.
The updated ODE equations involving the local population are
$$
\begin{align}
    \dot{S}^l_{f0} &= - (\beta_{00} I^l_{m0}+\beta_{10} I^l_{m1}\color{red}{+\beta_{20}I_m^x}) S^l_{f0} /N^l + \mu (S^l_{f0}/S^l) - \mu S^l_{f0} \\
    \dot{S}^l_{f1} &= - (\beta_{01} I^l_{m0}+\beta_{11} I^l_{m1}\color{red}{+\beta_{21}I_m^x}) S^l_{f1} /N^l + \mu (S^l_{f1}/S^l) - \mu S^l_{f1} \color{red}{+\nu S^x_{f}}\\
    \dot{S}^l_{m0} &= - (\beta_{00} I^l_{f0}+\beta_{10} I^l_{f1}\color{red}{+\beta_{20}I_f^x}) S^l_{m0} /N^l + \mu (S^l_{m0}/S^l) - \mu S^l_{m0} \\
    \dot{S}^l_{m1} &= - (\beta_{01} I^l_{f0}+\beta_{11} I^l_{f1}\color{red}{+\beta_{21}I_f^x}) S^l_{m1} /N^l + \mu (S^l_{m1}/S^l) - \mu S^l_{m1} \color{red}{+\nu S^x_{m}}\\
    \dot{I}^l_{f0} &= (\beta_{00} I^l_{m0}+\beta_{10} I^l_{m1}\color{red}{+\beta_{20}I_m^x} ) S^l_{f0} /N^l - \gamma I^l_{f0} - \mu I^l_{f0} \\
    \dot{I}^l_{f1} &= (\beta_{01} I^l_{m0}+\beta_{11} I^l_{m1}\color{red}{+\beta_{21}I_m^x} ) S^l_{f1} /N^l - \gamma I^l_{f1} - \mu I^l_{f1} \color{red}{+\nu I^x_{f}}\\
    \dot{I}^l_{m0} &= (\beta_{00} I^l_{f0}+\beta_{10} I^l_{f1}\color{red}{+\beta_{20}I_f^x} ) S^l_{m0} /N^l - \gamma I^l_{m0} - \mu I^l_{m0} \\
    \dot{I}^l_{m1} &= (\beta_{01} I^l_{f0}+\beta_{11} I^l_{f1}\color{red}{+\beta_{21}I_f^x} ) S^l_{m1} /N^l - \gamma I^l_{m1} - \mu I^l_{m1} \color{red}{+\nu I^x_{m}}\\
    \dot{T}^l_{f0} &= \gamma I^l_{f0} - \mu T^l_{f0} \\
    \dot{T}^l_{f1} &= \gamma I^l_{f1} - \mu T^l_{f1} \color{red}{+\nu T^x_{f}}\\
    \dot{T}^l_{m0} &= \gamma I^l_{m0} - \mu T^l_{m0} \\
    \dot{T}^l_{m1} &= \gamma I^l_{m1} - \mu T^l_{m1} \color{red}{+\nu T^x_{m}}\\ 
\end{align}
$$

The ODE model has $6$ states for the external population and $12$ states for the local population.

The ODE model has $12$ parameters, 
$$
(\beta, b_{00}, b_{01}, b_{10}, b_{11}, b_{20}, b_{21}, \xi, \gamma, \zeta, \mu, \nu),
$$
of which $1$ is specified by the constraint $b_{00}+b_{01}+b_{10}+b_{11}+b_{20}+b_{21}=0$, say $b_{11}=-(b_{01}+b_{10}+b_{11}+b_{20}+b_{21})$. 
We can further fix $\mu=1/60$ years$^{-1}$, $\gamma=1/2$ years$^{-1}$. 
With this, our ODE model has $9$ free parameters,
$$
(\beta, b_{00}, b_{01}, b_{10}, b_{20}, b_{21}, \xi, \zeta, \nu).
$$

`r kableExtra::text_spec("Discussion points:", color = "red")`

* the population size in the local pop is not constant 


## **Phylodynamic model**
For a phylodynamic analysis, we further reduce the number of states of the above
ODE model. We assume that

* the external population is at endemic equilibrium, with HIV prevalence of
  $0.1\%$ and of those $60\%$ are on treatment. We further assume that half the population is male. This
fixes 
$$
\begin{align}
	S^x_{f} &= 0.999 * 0.5 * N^x =: c^x_{Sf} N^x\\ 
	S^x_{m} &= 0.999 * 0.5 * N^x =: c^x_{Sm} N^x\\
	I^x_{f} &= 0.001 * 0.5* 0.4 * N^x =: c^x_{If} N^x\\ 
	I^x_{m} &= 0.001 * 0.5* 0.4 * N^x =: c^x_{Im} N^x\\ 
	T^x_{f} &= 0.001 * 0.5* 0.6 * N^x =: c^x_{Tf} N^x\\ 
	T^x_{m} &= 0.001 * 0.5* 0.6 * N^x =: c^x_{Tm} N^x\\
\end{align},
$$
and this should also fix the rates $\xi$, $\zeta$ to some value. 
The population size $N^x$ is still allowed to vary over time. 
* all individuals have a sequence sampled before ART start. The viral
  lineages can thus be in the states
$$
(I^l_{f0}, I^l_{f1}, I^l_{m0}, I^l_{m1}, I^x_{f}, I^x_{m}).
$$
* Gender is strongly informative to rule out transmission between two close
  females, and so we aim to keep gender in the model for the local population.
  However tracking lineages in the external population is expensive and I don t
  see how this could change estimates of $b_{00}$, $b_{01}$, $b_{10}$. We thus sum $I^x=I^x_{f} + I^x_{m}$ and reduce the state space of viral
  lineages to  
$$
(I^l_{f0}, I^l_{f1}, I^l_{m0}, I^l_{m1}, I^x),
$$
  and order the $5$-dimensional state space in this manner, i.e. the first dimension corresponds to
$I^l_{f0}$ etc.


Using standard notation, we define the populatin sizes in these $m=5$ demes backwards in time by
$$
\boldsymbol{Y}(t)= \big(  I^l_{f0}(t), I^l_{f1}(t), I^l_{m0}(t), I^l_{m1}(t), I^x(t) \big).
$$
The population sizes in the demes are further influenced by the dynamic
variables
$$
\boldsymbol{Y}^\prime(t)= \big(  S^l_{f0}(t), S^l_{f1}(t), S^l_{m0}(t), S^l_{m1}(t), S^x(t) \big).
$$
The birth matrix corresponding to the ODE model above is
$$
\boldsymbol{F}(t,\boldsymbol{Y}, \boldsymbol{Y}^\prime)= 
\left[\begin{array}
{rrrrr}
0 & 0 & \beta_{00}I^l_{m0}S^l_{f0}/N^l & \beta_{10}I^l_{m1}S^l_{f0}/N^l & \beta_{20} (I^x/2) S^l_{f0}/N^l  \\
0 & 0 & \beta_{01}I^l_{m0}S^l_{f1}/N^l & \beta_{11}I^l_{m1}S^l_{f1}/N^l & \beta_{20} (I^x/2) S^l_{f1}/N^l\\
\beta_{00}I^l_{f0}S^l_{m0}/N^l & \beta_{10}I^l_{f1}S^l_{m0}/N^l & 0 & 0 & \beta_{20} (I^x/2) S^l_{m0}/N^l\\
\beta_{01}I^l_{f0}S^l_{m1}/N^l & \beta_{11}I^l_{f1}S^l_{m1}/N^l & 0 & 0 & \beta_{20} (I^x/2) S^l_{m1}/N^l\\
0 & 0 & 0 & 0 & \xi I^xS^x/N^x\\
\end{array}\right],
$$
where we ommitted dependencies on time on the right hand side.
The entry $F_{ij}(t)$ describes the rate at which individuals in deme $i$ are generated by individuals in deme $j$.
The migration matrix corresponding to the ODE model above is
$$
\boldsymbol{G}(t,\boldsymbol{Y}, \boldsymbol{Y}^\prime)= 
\left[\begin{array}
{rrrrr}
0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & \nu (I^x/2)\\
0 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & \nu (I^x/2)\\
0 & 0 & 0 & 0 & 0\\
\end{array}\right].
$$
The entry $G_{ij}(t)$ describes the rate at which individuals in deme $i$ migrate from individuals in deme $j$.
The death vector corresponding to the ODE model above is
$$
\boldsymbol{\mu}(t,\boldsymbol{Y}, \boldsymbol{Y}^\prime)= 
\left[\begin{array}
{r}
\gamma I^l_{f0} \\
\gamma I^l_{f1} \\
\gamma I^l_{m0} \\
\gamma I^l_{m1} \\
\zeta I^x \\
\end{array}\right].
$$
With the above functions, the dynamics of $\boldsymbol{Y}$ can be computed by solving the ODE system
$$
\dot{Y}_j(t) = -\mu_k(t) + \sum_{i=1}^m ( F_{ij}(t) + G_{ij}(t) - G_{ji}(t) ).
$$

The phylodynamic model has 1 state for the external population ($N^x$) and 9 states for the local population 
($S^l_{f0}, S^l_{f1}, S^l_{m0}, S^l_{m1}$, $I^l_{f0}, I^l_{f1}, I^l_{m0}, I^l_{m1}$, $N^l$).
The phylodynamic model has parameters 
$$
(\beta, b_{00}, b_{01}, b_{10}, b_{11}, b_{20}, b_{21}, \xi, \gamma, \zeta, \nu),
$$
of which $1$ is specified by the constraint
$b_{00}+b_{01}+b_{10}+b_{11}+b_{20}+b_{21}=0$, say
$b_{11}=-(b_{01}+b_{10}+b_{11}+b_{20}+b_{21})$. We further fix $\gamma=1/2$ years$^{-1}$, and $\xi$, $\zeta$ to some value to obtain the pre-specified endemic equilibrium in the external
population. With this, our phylodynamic model has $7$ free parameters,
$$
(\beta, b_{00}, b_{01}, b_{10}, b_{20}, b_{21}, \nu).
$$

`r kableExtra::text_spec("Discussion points:", color = "red")`

* I am not clear on how to fix $\xi$, $\zeta$, given the pre-specified constraints and the migration dynamics.
* To solve the phylodynamic ODE of $\boldsymbol{Y}$, we need to know $\boldsymbol{Y}^\prime$ at time $t$. I don t understand how this is calculated, it seems to require other 
elements than the migration, birth and death functions, and other parameters such as $\mu$ above. 
 