---
title: "**data update 180706 -> 191011**"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include=TRUE, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(knitr)
require(kableExtra)
require(ggplot2)
```
```{r, include=FALSE, eval=FALSE, echo=FALSE, warning=FALSE, message=FALSE, tidy=TRUE}
require(rmarkdown)
setwd('~/git/hivclust/pkg/misc/vignettes') 
rmarkdown::render('seattle.191017.alignment.update.Rmd')
```

## **Data checks: sequence meta data**
Load two data sets: 
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE}
require(tidyverse)

home <- '~/Box Sync/OR_Work/Seattle'
infile.seqinfo.current <- file.path(home,'PHSKC-2018-07-09/sequences_meta.rds')
infile.seq.current <- file.path(home,'PHSKC-2018-07-09/Sequences.fas')
infile.seqinfo.update <- file.path(home,'PHSKC-2019_10_11/sequences_meta.rds')
infile.seq.update <- file.path(home,'PHSKC-2019_10_11/Sequences.fas')

sc <- readRDS(infile.seqinfo.current) %>%
		filter(type!='IN') %>%
		select(seqID, newnum, seqy, seqm) %>%	
		rename(cur.newnum=newnum, cur.seqy=seqy, cur.seqm=seqm)			
su <- readRDS(infile.seqinfo.update) %>%
		filter(type!='IN') %>%
		select(seqID, newnum, type, seqy, seqm) %>%
		rename(upd.newnum=newnum, upd.seqy=seqy, upd.seqm=seqm)
ss <- sc %>% full_join(su, by='seqID')
```

## **Test 1 (fail)**
I was expecting that every old seqID is part of the new data, but this
does not appear to be true:

```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE} 
ss %>% filter(is.na(upd.newnum))
```
For example:
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE} 
ss %>% filter(cur.newnum==876)
```

## **Test 2 (ok)**
I was expecting that the same seqIDs in the new and old data sets have the same
meta-data, which is true. 
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE} 
ss %>%
		filter(!is.na(upd.newnum) & !is.na(cur.newnum)) %>%			
		filter(cur.newnum!=upd.newnum | cur.seqy!=upd.seqy | cur.seqm!=upd.seqm ) 
```

## **Test 3 (fail)**
I was expecting that the same seqIDs in the new and old data sets have the same
sequence, which does not seem to be true.
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE} 
seq.ids <- ss %>% 
		filter(!is.na(upd.newnum) & !is.na(cur.newnum)) %>%
		pull(seqID)
seqc <- read.dna(infile.seq.current, format='fa')
sequ <- read.dna(infile.seq.update, format='fa')
test <- sapply(seq_along(seq.ids), function(i){
			cur <- paste(unlist(as.character(seqc[ seq.ids[i] ])), collapse='')
			upd <- paste(unlist(as.character(sequ[ seq.ids[i] ])), collapse='')
			cur==upd
		})
test <- tibble(seqID=seq.ids, identical=test)
test.fail <- test %>% filter(!identical)
test.fail 
```
For example:
```{r, include=TRUE, eval=TRUE, echo=TRUE, warning=FALSE, message=FALSE, tidy=TRUE} 
seq.name <- test.fail[1, ] %>% pull(seqID )
paste(unlist(as.character(seqc[ seq.name ])), collapse='')
paste(unlist(as.character(sequ[ seq.name ])), collapse='')
```

